The role of Similarity in native spoken word recognition

### 2025-01-13 manuscript/manuscript.qmd
Address NSG comments

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer*. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. Outrageous as these examples might sound, the reader may know a few cases in their own native language. This auditory illusion is common across languages and cultures [@dembeck2015oberflachenubersetzung; @efimova2018homophonic] and can feel quite real, and often inevitable. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"). Soramimi is a particular case of homophonic translation in which words or phrases that appear in the lyrics of a song in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning [@gasparov2006semen]. Homophonic translation also occurs outside the context of music. For instance, it is often intentionally employed by translators to preserve the meter and "sound structure" (e.g., rhymes) of an original text in the language of translation [e.g., @levick2016translating; @pilshchikov2016semiotics]^[The rationale behind this translation approach is to "sacrifice[s] fidelity to the image rather than the melodiousness of verse" [@briusov1911ot]. For instance, Edith Grossman's translation of *El Quijote* [@cervantes] present a similar case, in which formal aspects of the original text in Spanish are favoured to some degree over equivalence to English, even sometimes keeping Spanish words from the original text in the English translation (e.g., *Señor*, *ínsula*).]. Although homophonic translation is mostly seen as a perceptual curiosity or as a literary figure, little attention has been paid to homophonic translation as a spontanous psycholinguistic event.

@otake2007interlingual analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the TV show *Soramimi hour*, in which examples of Soramimi were presented the the audience for comedy purposes. These instances consisted of homophonic translations of English song lyrics to words and phrases in Japanese. The vast majority of the reported instances were phrases (96%) and few were single words (4%). There was relatively high variability in the degree to which phonetic features of the presented English lyrics were preserved, with some resulting Japanese words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. An analysis of the few instances of homophonic translations of single words revealed three main phonological processes that could explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /\textipa{kraI}/ to *kurai*, /\textipa{kuRai}/ [dark]), deletion (e.g., *go*, /\textipa{goU}/ to *go*, /go/ [go], and alternation (e.g., *low*, /\textipa{loU}/ to *rou*, /\textipa{Roo}/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic].

In the present work, we investigated the interplay between phonological similarity and phonological neighbourhood density to examine how listeners may activate lexical representations in their native language when listening to an unfamiliar language. Although homophonic translations do not necessarily take place with the preservation of meaning---which results in outrageous mistranslations like the aforementioned ones---homophonic translations sometimes may lead to correct translations. For example, imagine an English native listening to Dutch for the first time. This listener may encounter the word /\textipa{"pINgUIn}/ (*pinguin*), which sounds similar to its English translation /\textipa{"pENgwIn}/ (penguin). Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographic similarity of the former to the equivalent in their native language.

When comparing the phonemic transcription of the French to that of its Spanish translation /\textipa{"pweR.ta}/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /\textipa{K}/ and the open-mid back rounded vowel /\textipa{O}/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /\textipa{R}/ and /o/, as described by assimilation models of phoneme perception [@best2001discrimination; @best1988examination]. Although this phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish listener may still be able to activate the phonological representation of /\textipa{"pweR.ta}/ (*puerta*) from hearing the word-form /\textipa{"pOKt}/ (*porte*), ultimately allowing them to access the correct semantic representation. Native prosody also constrains the occurrence of homophonic translations. For instance, @kentner2015rhythmic presented in the auditory modality French-native and German-native participants with English songs with lyrics. Homophonic translations followed language-specific segmentation strategies. Most homophonic translations reported by German-native participants resulted from inserting phrase and word boundaries before stressed syllables (following the more frequent trochaic pattern in German). On the contrary, most homophonic translations reported by French-native participants resulted from inserting phrase and word boundaries after stressed syllables, following the more frequent iambic pattern in French.

@@

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer!* When Michael Jackson vocalised "ma-ma-coo-sah" in his song *Wanna Be Startin' Somethin'* (*Thriller*, 1982), some Dutch speakers understood *mama appelsap* [mama applejuice]. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. When Japanese speakers listen to the line "I want to hold your hand" in the homonym song by The Beatles (*Meet The Beatles*, 1964), they often mishear *Aho na hōnyōhan* [stupid urinator]. Outrageous as these examples might sound to speakers of other languages, many readers may know a few cases in their own native language. This auditory illusion can feel quite real, and often inevitable after the first time it happens. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"), after the *Soramimi Hour*---a section of the program *Tamori Club* hosted by comedian Morita Kazuyoshi (Tamori)---in which instances of this illusion were presented with comedy purposes. Soramimi is a particular case of *homophonic translation*: words or phrases in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning. This phenomenon extends beyond song lyrics, and may occur anytime when listening to non-native speech. For instance, when listening to "You know me?" (/ju: nəʊ mi:/) in English, Japanese-native listeners tend to understand *yunomi* (/jʊnomi/), which is a type of Japanese teacup. Overall, the occurence of Soramimi---and more generally, homophonic translations---suggests that speech in an unfamiliar language has the potential to activate lexical representations in the native language.

Although homophonic translations do not necessarily take place with the preservation of meaning---which often results in outrageous mistranslations like the ones aforementioned---homophonic translations sometimes may lead to the activation of the intended meaning (i.e., a correct translation). For example, imagine a Spanish native with no previous familiarity with any other language listening to French for the first time. This listener may encounter the word /pɔKt/ (*porte*), translation of door in French. When comparing the phonemic transcription of the French to that of its Spanish translation /pweɾta/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /ʁ/ and /o/, as described by assimilation models of phoneme perception [@best2001discrimination; @best1988examination]. Although this phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish listener may still be able to activate the phonological representation of /pweɾta}/ (*puerta*) from hearing the word-form /pɔʁt/ (*porte*), ultimately allowing them to access the correct semantic representation. Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographic similarity of the former to the equivalent in their native language.

Soramimi, and more generally homophonic translation, are popular and frequent phenomena, but the psycholinguistic literature on the topic is relatively scarce, and has mostly focused on the phonological variables involved in the occurence of Soramimi. For instance, @otake2007interlingual analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the TV show Soramimi hour. These instances consisted of homophonic translations of English song lyrics to Japanese words and phrases. The vast majority of the reported instances were phrases (96%) and a few, to single words (4%). The phonetic features of the presented English lyrics were preserved with relatively high variability, with some Japanese resulting words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. When analysing the few instances of homophonic translations of single words, the author identified three phonological processes that might explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /krʌɪ/ to *kurai*, /kuɾai/ [dark]), deletion (e.g., *go*, /goʊ/ to *go*, /go/ [go], and alternation (e.g., *low*, /loʊ/ to *rou*, /ɾoo/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic]. Native prosody also constrains the occurence of homophonic translations. For instance, @kentner2015rhythmic presented in the auditory modality French-native and German-native participants with English songs with lyrics. Homophonic translations followed language-specific segmentation strategies. Most homophonic translations reported by German-native participants resulted from inserting phrase and word boundaries *before* stressed syllables (following the more frequent trochaic patter in German). On the contrary, most homophonic translations reported by French-native participants resulted from inserting phrase and word boundaries *after* stressed syllables, following the more frequent iambic pattern in French.

While the previous studies have addressed the phonological mechanisms behind homophonic translation, little attention has been paid to the dynamics of lexical activation and competition that underlay this phenomenon. In the present work we investigated the interplay between phonological similarity and phonological neighbourhood density to examine how listeners of an unfamiliar language activate lexical representations in their native language. In particular, we studied how the presence of phonological neighbours in the native language lexicon affects the extent to which listeners are able to activate the correct translation of an auditorily presented word in an unfamiliar language.

Non-native speech perception is a vastly studied subject [@best1988examination; @miyawaki1975effect; @pallier2001influence; @weber2004lexical; @van2015learning; see @baese2020perception for review], but most of this research considers the case of participants with some degree of familiarity with the non-native language, characterising how bilinguals recognise, translate, and produce words as a function of their proficiency in each of their languages [e.g., @kroll1994category; @dijkstra2019multilink; @dijkstra2002architecture; @christoffels2006memory]. Even low-proficiency bilinguals are likely to have already acquired many form-meaning mappings in their lower proficiency language. Characterising the impact of phonological similarity on word processing is therefore challenging in bilinguals, due to the simultaneous contribution of phonological *and* conceptual links in the lexicon. By testing participants with *no* prior familiarity with the presented language (i.e., no proficiency), we ensured that no form-meaning associations were available for participants to correctly translate the unfamiliar words, if not by relying exclusively on phonological similarity.

In this study, we used a translation elicitation task in which participants listened to words from an unfamiliar language, and then tried to guess their translation in their native language. We will henceforth refer to the auditory-presented words heard by participants on each trial as presented words, and the correct translation for the presented words as target translations. We explored listeners’ reliance on phonological similarity to succeed in the task by manipulating the amount of phonological similarity between the presented words and their target translations. Since participants were unfamiliar with the presented language, they could only use phonological similarity between the presented language and their native language to successfully translate the words. We therefore predicted that participants’ performance should increase when the translation pairs are phonologically more similar. We also predicted that there is a minimum threshold of phonological similarity to be sufficient for translation.

We further investigated the effect of phonological competitors. Even if the presented word and its target translation share high phonological similarity, participants may still provide incorrect translations if the presented word also shares high phonological similarity with other words in the native language. Words with denser phonological neighbourhoods are recognised more slowly and less accurately than words with sparser phonological neighbourhoods [@dufour2003lexical; @goldinger1989priming; @hamburger1996phonological; @luce1990similarity; @luce1998recognizing]. This is especially true if such neighbours share higher phonological similarity with the presented word, or are lexically more frequent than the target translation [@luce1998recognizing; @dufour2010phonological]. Since the interference effects of phonological neighbourhood density has also been reported across languages [@weber2004lexical]. The presence of cross-linguistic pairs which are phonologically similar but differ in meaning (e.g., false friends) may act as distractors during lexical access, obstructing the selection of the appropriate target translation in the listener’s lexicon. For instance, @otwinowska2019more reported that false friends were disadvantaged relative to non-cognates by Polish second language learners, in contrast to cognates which were known better. It is therefore important to investigate the joint effect of both cognates and false friends when investigating the effect cross-linguistic phonological similarity has on word recognition in a foreign language. This could shed light on available strategies and challenges associated with the early stages of second language acquisition.

We hypothesised that the size of the facilitatory role of phonological similarity would be inversely proportional to the amount of higher-frequency phonological neighbours that the the presented word has in the target language. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as *bottom* instead of as its correct translation *button*, due to combined effect of the close phonological similarity between *botón* and *bottom* along with the high lexical frequency of *bottom* relative to *button*. To test this prediction, we developed a lexical frequency-dependent measure of cross-language phonological neighbourhood density, in which a neighbour is counted only if it is higher frequency and is one phoneme apart from the presented word. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that competitors in the native language affect recognition of non-native words during recognition of foreign speech.

We conducted a series of three experiments to test these predictions. In Experiment 1, we collected data from two groups of British English natives living in the United Kingdom. One group was presented with Catalan words, the other with Spanish words. We examined the extent to which participants were able to use the phonological similarity between the presented word (in Catalan or Spanish) and its target translation to provide accurate responses. In Experiment 2, we tested a group of (European) Spanish natives in the same task, who were presented with Catalan words. Catalan and Spanish are both Romance languages whose close typological distance is reflected in the fact that they share many cognates, while English is a Germanic language that shares considerably fewer cognates with Catalan and Spanish. By testing participants translating words from typologically close or distant languages, we expected to widen the range of the phonological similarity scores of the translation pairs involved in the experimental task, therefore allowing us to explore potential cross-language differences in participant’s performance. One unexpected finding was that participants in Experiments 1 and 2 were surprisingly good at translating a subset of words which had low phonological similarity with their correct translation. We were concerned that this may be caused by some prior knowledge of specific words by our participants, as some Spanish words are commonly seen in media or product labels, making them familiar even to monolingual speakers of English. In Experiment 3, we collected additional data from a new group of British English natives. The design was closely modelled after Experiment 1, except that after providing their response in each trial, participants reported whether they had previous knowledge of the presented word. In the final section of this paper, we present analyses on the joint data sets of Experiments 1 to 3.

@@

For each word in the Catalan and Spanish lists pair, we defined three predictors of interest: the lexical frequency of the correct English translation (*Frequency*), the phonological similarity between the presented word and their correct English translation (*Similarity*), and the presented word's number of cross-language phonological neighbours (*CLPN*) in English. *Frequency* was included as a nuance predictor, under the hypothesis that---keeping other predictors constant---participants would provide correct responses more often when the correct translations are high frequency words in their native language. Lexical frequencies of correct translations were extracted from SUBTLEX-UK [@van2014subtlex], and transformed to Zipf scores. Translations without a lexical frequency value were excluded from data analysis (`r stim_exc_freq$cat_eng$n` in the Catalan list, `r stim_exc_freq$spa_eng$n` in the Spanish list).

We calculated *Similarity*, our main predictor of interest, by computing the Levenshtein similarity between the X-SAMPA transcriptions of each pair of translations using the `stringdist` R package [@van2014stringdist]. The Levenshtein distance computes the edit distance between two character strings---in this case, two phonological transcriptions---by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. We divided this edit distance by the number of characters in the longest X-SAMPA transcription of the translation pair. This transformation accounts for the fact that Levenshtein distance tends to increase with the length of the transcriptions. For interpretability, we subtracted this proportion from one, so that values closer to one correspond to higher similarity between phonological transcriptions, and values closer to zero correspond to lower similarity between phonological transcriptions. For example, the *table* (`teIb@l`)-*mesa* (`mesa`) translation pair had `r scales::percent(stringdist::stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (`trEIn`)-*tren* (`tREn`) translation pair had `r scales::percent(stringdist::stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity.

For each Catalan and Spanish word, we calculated the number of *CLPN* by counting the number of English words with same or higher lexical frequency, and whose phonological transcription (in X-SAMPA format) differed in up to one phoneme from that of the presented Catalan or Spanish word. Lexical frequencies and phonological transcriptions were extracted from the multilingual database CLEARPOND [@marian2012clearpond]^[[Phonological transcriptions in CLEARPOND were generated from eSPEAK, [http://espeak.sourceforge.net/](http://espeak.sourceforge.net/)]]. @tbl-stimuli summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

We modelled the probability of participants guessing the correct translation of each presented word using a generalised multilevel Bayesian regression model with a Bernoulli logit link distribution. We included as fixed effects the intercept, the main effects of *Frequency*, *Similarity*, and *CLPN*, and the two-way interaction between *Similarity* and *CLPN*. We also included participant-level random intercepts and slopes for the main effects and the interaction. Eq. 1 shows a formal description of the model.

@@

To test the practical relevance of each predictor we followed @kruschke2018bayesian. We first specified a region of practical equivalence (ROPE) around zero ([-0.1, +0.1], in the logit scale). This area indicates the values of the regression coefficients that we considered as equivalent to zero. We then computed the 95% posterior credible intervals (CrI) of the regression coefficients of interest. Finally, we calculated the proportion of the 95% CrI that fell inside the ROPE. This proportion indicates the probability that the true value of the coefficient is equivalent to zero. All analyses were performed in R environment [@rcoreteam2013language]. We used the tidyverse family of R packages [@wickham2019welcome] to process data and to generate figures. We used the `brms` R package [@burkner2017brms] using the `cmdstanr` back-end to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for mode details on the models).

@@

```{r tbl-dataset}
#| label: tbl-dataset
sem <- function(x) mean(x) / (sqrt(length(x)))

@@

Stimuli were the same list of Catalan stimuli as in Experiment 1. Procedure and data analysis were identical as in Experiment 1.

@@

>>>>

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer*. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. Outrageous as these examples might sound, the reader may know a few cases in their own native language. This auditory illusion is common across languages and cultures and can feel quite real, often inevitable [@dembeck2015oberflachenubersetzung; @efimova2018homophonic]. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"). From a linguistic point of view, Soramimi is a particular case of homophonic translation: words or phrases that appear in the speech stream in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning [@gasparov2006semen]. Homophonic translation has mostly received attention as a perceptual curiosity, and as a literary figure, intentionally employed by translators to preserve the meter and "sound structure" (e.g., rhymes) of an original text in the language of translation [e.g., @levick2016translating; @pilshchikov2016semiotics]^[The rationale behind this translation approach is to "sacrifice[s] fidelity to the image rather than the melodiousness of verse" [@briusov1911ot]. For instance, Edith Grossman's translation of *El Quijote* [@cervantes] present a similar case, in which formal aspects of the original text in Spanish are favoured to some degree over equivalence to English, even sometimes keeping Spanish words from the original text in the English translation (e.g., *Señor*, *ínsula*).]. Little attention has been paid to homophonic translation as a psycholinguistic event.

In one of the few available studies concerning homophonic translation as a psycholinguistic phenomenon, @otake2007interlingual analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the Japanese TV show *Soramimi hour*, in which examples of Soramimi were presented to the audience for comedy purposes. These instances consisted of homophonic translations of English song lyrics to words and phrases in Japanese. The vast majority of the reported instances were phrases (96%) and few were single words (4%). There was relatively high variability in the degree to which phonetic features of the presented English lyrics were preserved, with some resulting Japanese words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. An analysis of the few instances of homophonic translations of single words revealed three main phonological processes that could explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /\textipa{kraI}/ to *kurai*, /\textipa{kuRai}/ [dark]), deletion (e.g., *go*, /\textipa{goU}/ to *go*, /go/ [go], and alternation (e.g., *low*, /\textipa{loU}/ to *rou*, /\textipa{Roo}/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic]. The phonological processes described by @otake2007interlingual do not take place in a vaccum, but rather in the context of a lexicon. As words or phrases in the native lexicon are activated by their acoustic similarity with the non-native speech stream, they trigger a series of mechanisms that result in the selection of one of the candidate word or phrase in the native language. In the present work, we investigated the interplay between phonological similarity and one of the main mechanisms involved in the dynamics of lexical selection: phonological neighbourhood density. The phonological neighbourhood density of a given word-form refers to the number of words-forms that differ in a single phoneme from the former. For instance, the English word *cat* hbelongs to a dense phonological neighbourhood, with 50 English phonological neighbours (e.g., hat, bat, cap), while the word *charger* belongs to a sparse phonological neighbourhood, with six neighbours (e.g., charged,larger, charter, charmer). Phonological neighbourhood density affects lexical processing. For instance, words from dense neighbourhoods are recognized more slowly and less accuractely in the auditory modality than words from sparse neighbourhoods [e.g., @luce1998recognizing; @vitevitch1998words; @dufour2010phonological; @vitevitch2006clustering]. In this study, we examined how listeners may activate lexical representations in their native language when listening to an unfamiliar language.

Although homophonic translations do not necessarily take place with the preservation of meaning---which results in outrageous mistranslations like the aforementioned ones---homophonic translations sometimes may lead to correct translations. For example, imagine an English native listening to Dutch for the first time. This listener may encounter the word /\textipa{"pINgUIn}/ (*pinguin*), which sounds similar to its English translation /\textipa{"pENgwIn}/ (penguin). Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and (orthographic similarity) of the former to the equivalent in their native language.

To explore the lexical mechanisms triggered by unfamiliar speech listening, we used a translation elicitation task in which participants listened to words from an unfamiliar language, and then tried to guess the words translations in their native language. We will henceforth refer to the auditory-presented words heard by participants on each trial as *presented words*, and the correct translation for the presented words as *target translations*. By manipulating the amount of phonological similarity between the presented words and their target translations, we explored listeners' reliance on phonological similarity to succeed in the task. Since participants were unfamiliar with the presented language, they could only use phonological similarity between the presented language and their native language to successfully translate the words. We therefore predicted that participants’ performance should increase when the translation pairs are phonologically more similar. We also predicted that there is a minimum threshold of phonological similarity to be sufficient for correct translation.

@@

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer*. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. Outrageous as these examples might sound, the reader may know a few cases in their own native language. This auditory illusion is common across languages and cultures and can feel quite real, often inevitable [@dembeck2015oberflachenubersetzung; @efimova2018homophonic]. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"). From a linguistic point of view, Soramimi is a particular case of homophonic translation: words or phrases that appear in the speech stream in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning [@gasparov2006semen]. Homophonic translation has mostly received attention as a perceptual curiosity, and as a literary figure, intentionally employed by translators to preserve the meter and "sound structure" (e.g., rhymes) of an original text in the language of translation [e.g., @levick2016translating; @pilshchikov2016semiotics]^[The rationale behind this translation approach is to "sacrifice[s] fidelity to the image rather than the melodiousness of verse" [@briusov1911ot]. For instance, Edith Grossman's translation of *El Quijote* [@cervantes] present a similar case, in which formal aspects of the original text in Spanish are favoured to some degree over equivalence to English, even sometimes keeping Spanish words from the original text in the English translation (e.g., *Señor*, *ínsula*).]. Little attention has been paid to homophonic translation as a psycholinguistic event.

In one of the few available studies concerning homophonic translation as a language processing phenomenon, @otake2007interlingual analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the TV show *Soramimi hour*, in which examples of Soramimi were presented the the audience for comedy purposes. These instances consisted of homophonic translations of English song lyrics to words and phrases in Japanese. The vast majority of the reported instances were phrases (96%) and few were single words (4%). There was relatively high variability in the degree to which phonetic features of the presented English lyrics were preserved, with some resulting Japanese words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. An analysis of the few instances of homophonic translations of single words revealed three main phonological processes that could explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /krʌɪ/ to *kurai*, /kuɾai/ [dark]), deletion (e.g., *go*, /goʊ/ to *go*, /go/ [go], and alternation (e.g., *low*, /loʊ/ to *rou*, /ɾoo/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic].

The phonological processes described by @otake2007interlingual do not take place in a vaccum, but rather in the context of a lexicon. As words or phrases in the native lexicon are activated by their acoustic similarity with the non-native speech stream, they trigger a series of mechanisms that result in the selection of one of the candidate word or phrase in the native language. In the present work, we investigated the interplay between phonological similarity and one of the main mechanisms involved in the dynamics of lexical selection: phonological neighbourhood density. In particular, we examined how listeners may activate lexical representations in their native language when listening to an unfamiliar language. Although homophonic translations do not necessarily take place with the preservation of meaning---which results in outrageous mistranslations like the aforementioned ones---homophonic translations sometimes may lead to correct translations. For example, imagine an English native listening to Dutch for the first time. This listener may encounter the word /ˈpɪŋ.guɪn/ (*pinguin*), which sounds similar to its English translation /ˈpɛŋ.ɡwɪn/ (penguin). Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographic similarity of the former to the equivalent in their native language.

When comparing the phonemic transcription of the French to that of its Spanish translation /ˈpweɾ.ta/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /ɾ/ and /o/, as described by assimilation models of phoneme perception [@best2001discrimination; @best1988examination]. Although this phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish listener may still be able to activate the phonological representation of /ˈpweɾ.ta/ (*puerta*) from hearing the word-form /ˈpɔʁt/ (*porte*), ultimately allowing them to access the correct semantic representation. Native prosody also constrains the occurrence of homophonic translations. For instance, @kentner2015rhythmic presented in the auditory modality French-native and German-native participants with English songs with lyrics. Homophonic translations followed language-specific segmentation strategies. Most homophonic translations reported by German-native participants resulted from inserting phrase and word boundaries before stressed syllables (following the more frequent trochaic pattern in German). On the contrary, most homophonic translations reported by French-native participants resulted from inserting phrase and word boundaries after stressed syllables, following the more frequent iambic pattern in French.

@@

For each word pair in the Catalan and Spanish lists, we defined three predictors of interest: the lexical frequency of the correct English translation (*Frequency*), the phonological similarity between the presented word and their correct English translation (*Similarity*), and the presented word's number of cross-language phonological neighbours (*CLPN*) in English. *Frequency* was included as a nuance predictor, under the hypothesis that---keeping other predictors constant---participants would provide correct responses more often when the correct translations are high frequency words in their native language. Lexical frequencies of correct translations were extracted from SUBTLEX-UK [@van2014subtlex], and transformed to Zipf scores. Translations without a lexical frequency value were excluded from data analysis (`r stim_exc_freq$cat_eng$n` in the Catalan list, `r stim_exc_freq$spa_eng$n` in the Spanish list).  *Similarity*, our main predictor of interest, was calculated as the Levenshtein similarity between the X-SAMPA transcriptions of each pair of translations using the `stringdist` R package [@van2014stringdist]. The Levenshtein distance computes the edit distance between two character strings---in this case, two phonological transcriptions---by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. We divided this edit distance by the number of characters in the longest X-SAMPA transcription of the translation pair. This transformation accounts for the fact that Levenshtein distance tends to increase with the length of the transcriptions. For interpretability, we subtracted this proportion from one, so that values closer to one correspond to higher similarity between phonological transcriptions, and values closer to zero correspond to lower similarity between phonological transcriptions. For example, the *table* (`teIb@l`)-*mesa* (`mesa`) translation pair had `r scales::percent(stringdist::stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (`trEIn`)-*tren* (`tREn`) translation pair had `r scales::percent(stringdist::stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. We calculated the number of *CLPN* for each Catalan and Spanish word by counting the number of English words with same or higher lexical frequency, and whose phonological transcription (in X-SAMPA format) differed in up to one phoneme from that of the presented Catalan or Spanish word. Lexical frequencies and phonological transcriptions were extracted from the multilingual database CLEARPOND [@marian2012clearpond]^[[Phonological transcriptions in CLEARPOND were generated from eSPEAK, [http://espeak.sourceforge.net/](http://espeak.sourceforge.net/)]]. @tbl-stimuli summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

We modelled the probability of participants guessing the correct translation of each presented word using a generalised multilevel Bayesian regression model with a Bernoulli logit link distribution. We included as fixed effects the intercept, the main effects of *Frequency*, *Similarity*, *CLPN*, and *Group* [sum-coded as `cat-ENG = -0`, `spa-ENG = +0.5`, @schad2020capitalize] and the three-way interaction between *Similarity*, *CLPN*, and *Group*. The predictor *Group* was included to account for any differences in task performance between English participants translating Catalan and Spanish words. We also included participant-level random intercepts and slopes for the main effects and the interaction. Eq. 1 shows a formal description of the model.

@@

To test the practical relevance of each predictor we followed @kruschke2018bayesian. We first specified a region of practical equivalence (ROPE) around zero ([-0.1, +0.1], in the logit scale). This area indicates the values of the regression coefficients that we considered as equivalent to zero. We then computed the 95% posterior credible intervals (CrI) of the regression coefficients of interest. Finally, we calculated the proportion of the 95% CrI that fell inside the ROPE. This proportion indicates the probability that the true value of the coefficient is equivalent to zero. All analyses were performed in R environment [@rcoreteam2013language]. We used the tidyverse family of R packages [@wickham2019welcome] to process data and to generate figures. We used the `brms` R package [@burkner2017brms] using the `cmdstanr` back-end to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for mode details on the models). Numeric predictors were standardised before entering the model by subtracting the mean and dividing by the standard deviation. Categorical predictors were sum-coded as `-0.5` and `+0.5` [@schad2020capitalize].

@@

```{r tbl-dataset}
#| label: tbl-dataset
#| tbl-cap: "Summary of participants' accuracy in the translation elicitation task across Experiments 1, 2, and 3."
sem <- function(x) mean(x) / (sqrt(length(x)))

@@

Stimuli were the same list of Catalan stimuli as in Experiment 1. Procedure and data analysis were identical as in Experiment 1, with the only exception that the statistical model did not include the *Group* predictor, given that only one group (`cat-SPA`) participated in this experiment.

@@

Experiment 3 was a conceptual replication of Experiment 1 in which we gathered additional information about English native participants' prior familiarity with the meaning of the Catalan and Spanish words presented. After each trial, participants reported whether they had prior knowledge about the word meaning, and how confident they were about such information in a numeric scale from 0 to 7. To control for the effect of participants' prior knowledge abbout form-meaning mappings, we excluded from the analyses those words in which participants reported such prior knowledge. We found equivalent results to those in Experiment 1, indicating that participants' surprisingly good performance in the translation elicitation task was not due to being familiar with the meaning of words they were presented with.
